---
title: "Homework 3"
output:
  html_document: default
  pdf_document: default
date: 'Deadline: 9 February, 12:00'
subtitle: 'Linguistic Data: Quantitative Analysis and Visualisation'
---

The solutions should be submitted via GitHub.  

## Part 1. A preliminary training  

*Do not use R (RStudio) to solve problems in Part 1. Answers won't be evaluated. *

### Problem 1

Look at the following histogram and answer the questions.

```{r, echo=FALSE}
set.seed(1234)
x <- rnorm(20, mean=3, sd=6)
hist(x, col="tomato", main="")
```

a. What is the proportion of values in the sample that exceed $5$? Explain your answer.
The proportion of values is1/4 because the total number of values  in the sample seems to be 20 (1+10+4+4+1=20) and 5 of them are plotted on the right of x = 5.  This means the number ofvalues that exceed 5 is (4+1=5). Thus, the propotion of values is 5/20 = 1/4.
b. Indicate the interval where the median of this sample can lie. Explain your answer.
Ithe median of this sample can lie in the interval (-5:0) because the median is calculated as the average  of the tenth and the eleventh values of the ordered sample and they lie in the interval (-5:0).
c. How the histogram will change if we add an element $7$ to the sample? Explain your answer.
 The frequency of values in the interval (5:10)  will get higher as well as the corresponding bar.
### Problem 2

a. Look at the histograms of two samples. They illustrate the distribution of normalized average reaction time to frequent words (in ms) in two groups of people.

```{r, echo=FALSE, fig.height=3}
set.seed(1234)
sample1 <- rnorm(100, 500, 50)
sample2 <- rnorm(100, 500, 20)
hist(sample1, col="lightblue", xlim=c(100, 900))
```

```{r, echo=FALSE, fig.height=3}
hist(sample2, col="lightgreen", xlim=c(100, 900))
```

Which of the samples has a larger variance? Explain your answer.
The first sample has larger variance because the  values in  sample1 differ more and are further apart (occupy broader interval) from what seems to be a mean of the distribution.

\newpage

b. Look at the histograms of two samples.

```{r, echo=FALSE, fig.height=8}
set.seed(12)
sample1 <- rnorm(100, 3, 5)
sample2 <- rnorm(100, 10, 0.8)
par(mfrow=c(2,2))
hist(sample1, col="lightblue", xlim=c(-40, 40))
hist(sample2, col="lightgreen", xlim=c(5, 15))
```

Which of the samples has a larger variance? Explain your answer.
The first sample has larger variance because the  values in  sample1 differ more and are further apart (occupy broader interval) from what seems to be a mean of the distribution.
## Part 2

*Do not use R (RStudio) to solve problems in Part 2. Answers for problem 3 will be evaluated. Please paste YES or NO into (empty) code blocks and explain you answer below the block.*

### Problem 3

Below is the histogram of the number of mistakes students made while writing an examination essay in English. Look at the histogram and answer the questions.

```{r, echo=FALSE, fig.height=4}
set.seed(2)
mistakes <- rbeta(1000, 2, 7)*100
hist(mistakes, breaks=60, col= "deepskyblue", ylim=c(0, 40))
```

### 3.1
Is it true that $50$% students made more than $35$ mistakes?
```
No
```
Explain your answer below:
Area right of 35 is less than half. Thus, the distribution shows that students who made more than 35 mistakes is less than 50%.

### 3.2
Is it true that most students made no more than $10$ mistakes?
```
NO
```

Explain your answer below: 
Area left of 10 is small. The value of 10 lies on the mistakes axis somewhere in the middle between 0 and 20, and it seems that there are more values on the right of 10.

### 3.3
Which of the following values is closer to be the median of `mistakes`: $10$, $20$, $30$, $40$?
```
20
```
Explain your answer below: 20 splits the total area in half.

### Problem 4. Exact binomial test

The null hypothesis is that $p=0$ (i.e. no success is possible). In a dataset, there is only one success out of 1 000 000 observations. Will you reject the null hypothesis? 
```
YES
```
Explain your answer below:
Success is possible.

## Part 3

*Use R (RStudio) to solve problems in Part 3. Your answers will be evaluated. Please paste R code into R code blocks and explain you answer below the block, if needed. *

### Problem 5 

Here is a sample of respondents' age:  

$44$, $50$, $42$, $64$, $66$, $42$, $72$, $56$, $72$, $54$, $46$, $48$, $48$, $52$, $50$, $66$, $84$.

### 5.1
Arrange them in a vector and call it `age`. 
```{r}
age <- c(44, 50, 42, 64, 66, 42, 72, 56, 72, 54, 46,
        48, 48, 52, 50, 66, 84)
```

### 5.2
Examine the type of `age` variable (numeric, character, etc).
```{r}
class(age) 
```

### 5.3
Plot the histogram of the vector `age` with $5$ bins. Change its color to any you want. (Use either R basic or ggplot2 style for plotting.)
```{r}
hist(age, breaks=5, col="green")
```


### Problem 6

Here is a series of words:  
*pie, bar, bar, pie, pie, bar, bar, chart*.

### 6.1
Arrange elements above in a vector and call it `words` 
```{r}
words = c("pie", "bar", "bar", "pie", "pie", "bar", "bar", "chart")
```

### 6.2
Calculate the relative frequences of values in `words` measured in percent.
```{r}
table(words)/length(words) * 100
```

## Problem 7. Position of verbs in verses

The dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv) is based on texts written in the 1820s and 1920s (Corpus of Russian Poetry of the Russian National Corpus). Authors collected only one line per author to keep observations as independent from each other as possible. 

Variables:  
- Decade — decade of creation: 1820s, 1920s.  
- RhymedNwords — the number of words in the rhyming position (usually one word, but there are two words in cases such as _вина бы_ 'I would like to get) wine' (which is rhymed with _жабы_ 'toad', see http://russian-poetry.ru/Poem.php?PoemId=18261)).  
- RhymedNsyl — the number of syllables in the rhyming position.  
- UPoS — part of speech of the last word. 
- LineText — a sampled verse.  
- Author — the author of the text.  

Can we decide that in verses written in 1920s, verbs in the rhyming posision are used differently (more often or less often) than expected for verbs in general? 

Let's assume that the probability for verbs to be used in any position ('in general') is 17% (according to [](http://www.ruscorpora.ru/new/corpora-stat.html) ).

### 7.1 State hypothesis

What is your null hypothesis $H_0$ and what is the alternative hypothesis $H_1$?
```
H0: In verses written in 1920s, verbs in the rhyming position are used in the same way verbs in general  are used with a probability of 0,17.
H1: In verses written in 1920s, verbs in the rhyming position are used differently (more often or less often) than expected for verbs in general (the probability is not 0,17).
```

### 7.2

Read the dataset [“The last words in verses”](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv). Filter out the relevant observations from 1920s, calculate the number of verbs observed in the sample, and the sample size. 
```{r}
data= read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/poetry_last_in_lines.csv", sep="\t")
nrow(data)
```
Observations from 1920s:
```{r}
twentieth_century_observations <- df[df$Decade=='1920s',]
```

The number of verbs observed in the sample:
```{r}
is.verb = nrow(twentieth_century_observations[twentieth_century_observations$UPoS=='VERB',])
is.verb
```

The sample size:
```{r}
sample_size = nrow(twentieth_century_observations)
sample_size
```
### 7.3

Use an exact binomial test to calculate p-value. 
```{r}
is.verb = data$UPoS[data$Decade=="1920s"] == "VERB"
binom.test(sum(is.verb), length(is.verb), p=0.17, alternative = "two.sided")
```

### 7.4 Interpret results

Give your interpretation of obtained p-value. Answer the initial question: Can we decide that in verses written in 1920s, verbs are used in the rhyming posision more often or less often than expected?
```
NO, we cannot. It is because the resulting p-value is  much bigger than alpha (0.05). So, it is not small enough. Hence, we can't reject the null hypothesis
```

### 7.5 
*(A bonus problem, extra points in evaluation)*.
Repeat 2.3 for verses written in the 1820s.
```{r}
is.verb = data$UPoS[data$Decade=="1820s"] == "VERB"
binom.test(sum(is.verb), length(is.verb), 0.17, alternative = "two.sided")
```

Write down your general conlusions about data provided for both 1920s and 1820s data.
```
Reuslting outcomes from both data samples indicate that verbs are  not  used differently  during these two periods of time. They also show that there is no big difference in the use of verbs in general as comparedto their use  in the rhyming position. Thus, the null hypothesis can't be rejected and we can't say that .

### Problem 8. One-sample t-test
Using Icelandic data on vowel duration from seminar [Link](https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv) test the null hypothesis that the population mean of vowel duration in speaker `shg05` equals 73 (ms).
To perform a one-sample t-test, you can use the following example of R code: 
```
t.test(sample, mu = 7725) # mu is a population mean
```

### 8.1
Write down a two-tailed alternative hypothesis.
```
H1: Mean vowel duration of speaker shg05 does not differ from 73ms
```

### 8.2
Perform a one-sample t-test. 
```{r}
data = read.csv("https://raw.githubusercontent.com/LingData2019/LingData2020/master/data/icelandic.csv")
vowel.dur = data$vowel.dur[data$speaker=="shg05"]
t.test(vowel.dur, mu = 73) # mu is a population mean
```

### 8.3
Interpret results. 
```
The true mean could be 73ms. We cannot reject this with 5% significance.
```

#### Supplementary reading
Use of exact binomial test in lingiistic research:  

* Gries, Stefan Th. "Phonological similarity in multi-word units." Cognitive Linguistics 22.3 (2011): 491-510. [Link](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.186.7412&rep=rep1&type=pdf)  
Stefan Gries proves that alliteration is observed in multi-word expressions more often than in general.

* Harald Bayen (2008: 51-52) evaluates the probability of observing exactly one occurrence of the word _hare_ in the corpus sample of 1 mln words given its estimated frequency of 8.23 words per million according to the SELEX frequency database.

On measures of central tendency: 

* Levshina 2015, Chapter 3 (p. 48); Gries 2009, Chapter 1.3 (p. 116). 

On t-test: 

* Gries 2009, Chapter 3 (p. 198). 